---
title: "AdverseDrugReactions"
author: "Thirupathi Pattipaka"
date: "7/5/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#setwd("/Users/thirupathipattipaka/OtherProjects/FDA_drug_adverse_event")
```

## Introduction

The U.S. Food and Drug Administration (FDA) regulates over-the- counter and prescription drugs in the United States, including biological therapeutics and generic drugs. This work covers more than just medicines. For example, fluoride toothpaste, antiperspirants, dandruff shampoos and sunscreens are all considered drugs.

<br>

An adverse event is submitted to the FDA to report any undesirable experience associated with the use of a medical product in a patient. For drugs, this includes serious drug side effects, product use errors, product quality problems, and therapeutic failures for prescription or over-the- counter medicines and medicines administered to hospital patients or at outpatient infusion centers.

<br>

The FDA's database of adverse event reports is made available through a web API at https://open.fda.gov/drug/event/. Each report contains general information about the report, patient information, a list of the drugs that the patient is taking, and a list of the patient reactions. It is possible to use these data in many ways: your brief is to explore these data and to see what might be learned from them.

The analysis is aimed to understand whether different adverse events reported in different countries?

## Loading list of libraries

```{r,echo=FALSE}
library(jsonlite)
library(dplyr)
library(stringi)
library(ggplot2)
library(downloader)
library(lubridate)
library(tidyr)
library(gtools)
library(knitr)
```

## Downloading data from openfda website(JSON files)

I am downloading drug adverse event data <https://open.fda.gov/drug/event/reference/>. The openFDA drug adverse event API returns data from the FDA Adverse Event Reporting System (FAERS), a database that contains information on adverse event and medication error reports submitted to FDA.


There are 87 data files and we will use a for loop to download all of them. Then, we will concatenate them.

```{r,echo=FALSE}
for (i in 80:87) {
  if (i<10){
    url = paste0("https://download.open.fda.gov/drug/event/all_other/drug-event-000",i,"-of-0087.json.zip")
  }
  else{
    url = paste0("https://download.open.fda.gov/drug/event/all_other/drug-event-00",i,"-of-0087.json.zip")
  }
  download(url,dest="json.zip")
  unzip ("json.zip")
  }
```

To check whether all files are downloaded
```{r}
dir()
```

To create a character vector of the all json files
```{r}
filenames <- list.files(pattern="*.json", full.names=TRUE)
filenames
```

## Understanding the data
To read JSON data, I am using the fromJSON function from the jsonlite package.
Let's read in one of the data files and print the content.
```{r}
drug1 = fromJSON(filenames[1])
```

```{r}
names(drug1)
```

what meta contains
```{r}
drug1$meta
```

what results contains
```{r}
results=drug1$results
class(results)
```


```{r}
glimpse(results)
```

As we can above, results contains 27 variables, out of those patients, sender,primarysource,reportduplicate and receiver are dataframes and others are character variables

```{r}
glimpse(results$patient)
```
The patient dataframe contains eight variables including with dataframe called summary and other variables are lists

To explore further about reaction list
```{r,echo=FALSE}
glimpse(results$patient$reaction[1])
```

The reaction contains 15000 lists, each patient with one list. Each list contains reactionmeddrapt variable which is included with the list of adverse reactions

Understand drug list in patient

```{r}
glimpse(results$patient$drug[1])
```
The list drug contains variables like medicinalproduct, drugindication and other dataframe openfda which contains variables like manufacturer_name,generic_name,brand_name etc.,

<br>
To understand primary source

```{r}
glimpse(results$primarysource)
```
<br>
All three variables in the primary source are character variables

```{r}
glimpse(results$reportduplicate)
```
<br>
All two columns in the reportduplicate are character variables

```{r}
glimpse(results$receiver)
```
All two columns in the receiver are also character variables

## Merging the data

Now, let's select certain variables from each data file and merge them and see adverse events reported in different countries
As a starting point, I will select drugname, drugindication, occurcountry, Adverse event, reportercountry, primarysourcecountry, etc., 

```{r}
#drugname
drugname <-  NULL

for (i in 1:15000) { 
  drugname <- rbind(drugname, results$patient$drug[[i]][["medicinalproduct"]] %>% paste( collapse=","))
}


#drugindication
drugindication <- NULL

for (i in 1:15000) { 
  drugindication <- rbind(drugindication, results$patient$drug[[i]][["drugindication"]] %>% paste( collapse=","))
}

AE <- NULL
for (i in 1:15000) { 
  AE <- rbind (AE, results$patient$reaction[[i]][["reactionmeddrapt"]] %>%  unlist() %>% paste(collapse=","))
  }

df <-  cbind(results[,!(names(results) %in% c("patient","sender","receiver","primarysource","reportduplicate"))],
             results$patient[,!(names(results$patient) %in% c("reaction","drug","patientdeath","summary"))],
             results$primarysource,
             drugname = drugname,
             drugindication= drugindication,
             AE = AE)

```
<br>
Merging all files

```{r,warnings=FALSE}

for(j in 2:8){                      # read all the rest
      
      tmp =fromJSON(filenames[j])$results
      #drugname
      drugname <-  NULL

      for (i in 1:nrow(tmp)) { 
          drugname <- rbind(drugname, tmp$patient$drug[[i]][["medicinalproduct"]] %>% paste( collapse=","))
      }


      #drugindication
      drugindication <- NULL

      for (i in 1:nrow(tmp)) { 
          drugindication <- rbind(drugindication, tmp$patient$drug[[i]][["drugindication"]] %>% paste( collapse=","))
      }
      
      AE <- NULL
      for (i in 1:nrow(tmp)) { 
          AE <- rbind (AE, tmp$patient$reaction[[i]][["reactionmeddrapt"]] %>%  unlist() %>% paste(collapse=","))
       }
      #df.new <-  cbind(tmp,tmp$patient,tmp$primarysource,drugname = drugname,drugindication= drugindication)
      df_new <-  cbind(tmp[,!(names(tmp) %in% c("patient","sender","receiver","primarysource","reportduplicate"))],
                       tmp$patient[,!(names(tmp$patient) %in% c("reaction","drug","patientdeath","summary"))],
                       tmp$primarysource,
                       drugname = drugname,
                       drugindication= drugindication,
                       AE = AE)

   df=smartbind(df, df_new)  # concatenate all
}
```

Final dataset for further processing

```{r}
glimpse(df)
```

```{r}
df = mutate(df,receive_date =ymd(receivedate))
df%>% group_by(receive_date) %>% summarize(Total=n()) %>% ggplot(aes(x=receive_date,y=Total)) + geom_line()
```
```{r}
#categorization of source country
df_ <- df %>%  select(safetyreportid, occurcountry,reportercountry, AE,drugname,patientonsetage,serious,primarysourcecountry,drugindication)
df_$occurcountry <- ifelse(is.na(df_$occurcountry),df_$primarysourcecountry,df_$occurcountry)
df_ <- df_ %>% mutate(occurcountry_g = case_when(.$occurcountry=='US'~"United states",
                                               .$occurcountry=='JP'~"Japan",
                                               .$occurcountry=='GB'~"Great Britain",
                                               .$occurcountry=='FR'~"France",
                                               .$occurcountry =='CA'~"Canada",
                                               .$occurcountry =='DE'~"Germany"
                                                ))
df_$occurcountry_g <- ifelse(is.na(df_$occurcountry_g),"Other",df_$occurcountry_g)

df2 <- df_ %>%
       mutate(AE2 = strsplit(as.character(AE), ",")) %>%
       unnest(AE2)
```

## Adverse events by country
```{r,fig.width=15}
df2%>% group_by(occurcountry) %>%summarize(Total=n()) %>% filter(Total >10) %>% ggplot(aes(x=occurcountry,y=Total))+ geom_bar(stat="identity")
df2%>% group_by(occurcountry) %>%summarize(Total=n()) %>% filter(Total >10) %>%  ungroup() %>%
    arrange(desc(Total)) %>% kable()
```


Countries with more than two adverse events reported in each report

```{r,fig.width=15}
df2%>% group_by(occurcountry) %>% summarize(Total=n()/n_distinct(safetyreportid)) %>% filter(Total >2) %>% ggplot(aes(x=occurcountry,y=Total))+ geom_bar(stat="identity")
```


## Top 100 Adverse drug reactions
```{r}

df2 %>% select(AE2) %>% table() %>% sort(... = 2,decreasing = TRUE) %>% as.data.frame() %>% head(100) %>%  kable()
```

Adverse events by country
```{r}
df2 %>% 
  group_by(AE2,occurcountry_g) %>%
  summarise(Total = n()) %>% 
  mutate(Total_events= sum(Total)) %>% spread(occurcountry_g, Total) %>% arrange(-Total_events) %>% head(100) %>%  
  kable(col.names=c("Adverse event","Total number of events","Canada","France","Germany","Great Britain","Japan","Other","USA"))
```
<br>

BY occur country groups
```{r}
df2%>% group_by(occurcountry_g) %>%summarize(Total=n()) %>%  mutate(Percentage=100*Total/sum(Total)) %>% kable(col.names=c("Country name","Total number of safety reports","Percentage"),digits=1)
```

<br>
Percentage of events in each country
```{r}
df2 %>% 
  group_by(AE2,occurcountry_g) %>%
  summarise(Total = n()) %>% mutate(freq=100*Total/sum(Total)) %>% 
  mutate(Total_events= sum(Total)) %>% select(AE2,occurcountry_g,Total_events,freq) %>% spread(occurcountry_g, freq) %>% arrange(-Total_events) %>% head(100) %>% 
  kable(col.names=c("Adverse event","Total number of events","Canada","France","Germany","Great Britain","Japan","Other","USA"),digits=1)
```


## Top drugs
```{r}
df %>% select(drugname) %>% table() %>% sort(... = 2,decreasing = TRUE) %>% as.data.frame() %>% head(100) %>%  
  kable(col.names=c("Drug name","Number of safety records"))
```

# Top diseases
```{r}
df %>% select(drugindication) %>% table() %>% sort(... = 2,decreasing = TRUE) %>% as.data.frame() %>% head(100) %>%  
  kable(col.names=c("Drug indication","Number of safety reports"))
```




